---
title: "The HTA Verification Framework"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The HTA Verification Framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(testhta)
library(testthat)
```

## Overview

Health Technology Assessment (HTA) models inform critical healthcare coverage, pricing, and clinical policy decisions. However, complex decision-analytic models are prone to structural bugs, indexing errors, and hidden parameter miswirings. 

The **HTA Verification Framework** in `testhta` provides a software-engineering-inspired methodology for formal model verification. Based on seminal frameworks by **Tappenden & Chilcott (2014)** and **Elbasha & Dasbach (2017)**, the framework applies automated "black-box" unit tests to verify that model code strictly behaves according to expected mathematical and economic boundary conditions.

---

## Verification Test Registry (T01 – T16)

The framework organizes 16 automated verification test cases into five distinct logical categories:

### 1. QALY Bounding Tests
These tests assert quality-adjusted life year (QALY) calculations under extreme discount rates and boundary utility weights:
- **T01 (Extreme Discounting 100%)**: Setting `discount_rate = 1` zeroes out future cycles, yielding total QALYs = 0.
- **T02 (Upper Bound QALYs)**: Setting state utilities = 1 with zero discounting (`discount_rate = 0`) forces QALYs to equal total undiscounted Life Expectancy (LE).
- **T06 (Utility-1 Scaling)**: Setting state utilities = 1 yields QALY gains equal to Life-Years (LYGs).
- **T07 (Discount Directionality)**: Standard discounting (e.g. 3.5%) strictly reduces QALYs compared to undiscounted runs.
- **T08 (Infinite Discounting)**: Infinite discount rates (`discount_rate = 1000`) drive future QALY gains to zero.

### 2. Cost Bounding & Sensitivity Tests
These tests check cost accumulation logic and parameter sensitivity:
- **T06 (Zero Cost Bounding)**: Setting all health state occupancy, transition, and intervention costs to 0 must output total costs of 0.
- **T09 (Zero Intervention Cost)**: Removing intervention costs strictly reduces the Incremental Cost-Effectiveness Ratio (ICER).
- **T10 (Increased Intervention Cost)**: Increasing intervention cost strictly increases the ICER.
- **T11 (Zero Cost Discounting)**: A zero cost discount rate ensures discounted costs equal undiscounted costs.
- **T12 (Infinite Cost Discounting)**: An infinite cost discount rate reduces future costs to zero.

### 3. Population & Life Expectancy Bounding Tests
These tests verify cohort flow across state transitions:
- **T05 (Absolute Mortality)**: Setting death transition probability = 1 forces the cohort to die immediately, yielding Life Expectancy = 1 cycle.
- **T04 (Zero Mortality)**: Setting death transition probability = 0 ensures full cohort survival, yielding Life Expectancy equal to the full time horizon (`n_cycles`).

### 4. Treatment Arm Comparison & Symmetry Tests
These tests verify parameter wiring and symmetry across model arms:
- **T14 (Arm Parameter Equivalence)**: Setting all treatment-specific parameters equal across arms ($\theta_a = \theta_b$) yields identical costs and QALYs (zero incremental outcomes).
- **T15 (Parameter Sensitivity Sweep)**: Modifying an individual parameter for a treatment arm ($\theta_a \neq \theta_b$) must alter the ICER, confirming parameter wiring.
- **T16 (Arm Parameter Swapping)**: Swapping parameter vectors between arms ($\theta_a \leftrightarrow \theta_b$) swaps output costs and QALYs between those arms.

### 5. Probabilistic Sensitivity Analysis (PSA) Sampling Tests
These tests check parameter distribution limits before running Monte Carlo simulations:
- **T13 (Distribution Bounding)**: Verifies that PSA parameter draws strictly lie within theoretical domain limits (e.g., Beta utilities/probabilities $\in [0, 1]$, Lognormal costs $> 0$).

---

## Verification Test Summary Matrix

| Group | Test Case ID | Boundary / Verification Condition | Expected Behavior | Verification Helper / Code |
| :--- | :--- | :--- | :--- | :--- |
| **QALY Bounding** | **T01** | Set discount rate to `100%` (`discount_rate = 1`) | QALYs must equal `0` | `check_model_qalys(0, ...)` |
| **QALY Bounding** | **T02** | Set utilities to `1`, discount to `0` | QALYs must equal Life Expectancy | `expect_equal(get_qalys(r), get_le(r))` |
| **QALY Bounding** | **T06** | Set utilities for living states to `1` | QALY gains equal Life-Years (LYGs) | `expect_equal(get_qalys(r), r$total_LYs)` |
| **QALY Bounding** | **T07** | Standard discounting vs. undiscounted | Discounted QALYs strictly less than undiscounted | `compare_model_runs(..., expect_lt)` |
| **QALY Bounding** | **T08** | Infinite discount rate (`discount_rate = 1000`) | QALYs tend to zero | `check_model_qalys(0, ...)` |
| **Cost Bounding** | **T06** | Set all cost parameters to `0` | Total costs must equal `0` | `check_model_costs(0, ...)` |
| **Cost Bounding** | **T09** | Set intervention costs to `0` | ICER is reduced | `compare_model_runs(get_icer, expect_lt)` |
| **Cost Bounding** | **T10** | Increase intervention costs | ICER is increased | `compare_model_runs(get_icer, expect_gt)` |
| **Cost Bounding** | **T11** | Zero cost discount rate (`discount_rate = 0`) | Discounted costs equal undiscounted costs | `compare_model_runs(get_costs, expect_equal)` |
| **Cost Bounding** | **T12** | Infinite cost discount rate (`discount_rate = 1000`) | Costs tend to zero | `check_model_costs(0, ...)` |
| **Population / LE** | **T05** | Transition to death state set to `1` | Life Expectancy must equal `1` cycle | `check_model_le(1, ...)` |
| **Population / LE** | **T04** | Transition to death state set to `0` | Life Expectancy equals `n_cycles` | `check_model_le(n_cycles, ...)` |
| **Treatment Arm** | **T14** | Set all treatment parameters equal ($\theta_a = \theta_b$) | Costs and QALYs equal across arms | `expect_equal(get_incremental_costs(r), 0)` |
| **Treatment Arm** | **T15** | Amend individual model parameter ($\theta_a \neq \theta_b$) | ICER is modified (verifies parameter sensitivity) | `compare_model_runs(get_icer, expect_false)` |
| **Treatment Arm** | **T16** | Swap treatment-specific parameters ($\theta_a \leftrightarrow \theta_b$) | QALYs and costs swap between arms | `expect_equal(costs_swapped["arm1"], costs_base["arm2"])` |
| **PSA / Sampling** | **T13** | Draw $n$ samples from parameter distributions | All drawn values satisfy domain bounds (e.g. Beta $\in [0,1]$, Lognormal $>0$) | `expect_true(all(u >= 0 & u <= 1))` |

---

## References

1. **Tappenden, P., & Chilcott, J. B.** (2014). Avoiding and Identifying Errors and Other Threats to the Credibility of Health Economic Models. *PharmacoEconomics*, 32, 967–979. [https://doi.org/10.1007/s40273-014-0186-2](https://doi.org/10.1007/s40273-014-0186-2)
2. **Elbasha, E. H., & Dasbach, E. J.** (2017). Verification of Decision-Analytic Models for Health Economic Evaluations: An Overview. *PharmacoEconomics*, 35, 673–683. [https://doi.org/10.1007/s40273-017-0508-2](https://doi.org/10.1007/s40273-017-0508-2)
3. **Alarid-Escudero, F., et al.** (2019). A Need for Change! A Coding Framework for Improving Transparency in Decision Modeling. *PharmacoEconomics*, 37(11), 1329–1339. [https://doi.org/10.1007/s40273-019-00837-x](https://doi.org/10.1007/s40273-019-00837-x)
